\newpage
\section{The Strategic Setting}

\subsection{Causes of the Conflict}

%\begin{quote}
%  What do we mean by the Revolution? The war? That was no part of the Revolution;
%  it was only an effect and consequence of it. The Revolution was in the minds of
%  the people, and this was effected, from 1760 to 1775, in the course of
%  fifteen years before a drop of blood was shed at Lexington. The records of
%  thirteen legislatures, the pamphlets, newspapers in all the colonies, ought
%  to be consulted during that period to ascertain the steps by which the public
%  opinion was enlightened and informed concerning the authority of Parliament
%  over the colonies.\footnote{John Adams to Jefferson, 1815}
%\end{quote}

\begin{quote}
  ``John Adams felt strongly that Americans tended to confuse the American 
  Revolution with the Revolutionary War. The war, he wrote to Thomas Jefferson,
  `was no part of the Revolution. It was only an effect and consequence of it.'
  Adams, well aware that he was presenting and unconventional view, went on to
  state that the Revolution had taken place in the minds of the people from 1760
  to 1775, `before a drop of blood was spilled at
  Lexington.'\,''\cite[15]{dupuy_1977}
%  \footnote{``What do we mean by the
%  Revolution? The war? That was no part of the Revolution; it was only an effect
%  and consequence of it. The Revolution was in the minds of the people, and this
%  was effected, from 1760 to 1775, in the course of fifteen years before a drop
%  of blood was shed at Lexington. The records of thirteen legislatures, the
%  pamphlets, newspapers in all the colonies, ought to be consulted during that
%  period to ascertain the steps by which the public opinion was enlightened and
%  informed concerning the authority of Parliament over the colonies.'' (John Adams
%  to Jefferson, 1815)}
\end{quote}

There are many factors of politics, economics and policy that all affected the
colonies break away from Britain. One analysis of the American Revolution sees it
as a continuation of the Seven Years War. 
%Another focuses on its nature as a revolt against tightening of mercantilism
%around the colonies, who were slowly gravitating towards ideas of trade codified
%by Adam Smith's ``Wealth of Nations,'' published in 1776.
The popular analysis lies in a revolt against laws seen as unjust, levied
without representation. This section will address in turn the previous wars,
mercantilism, the influence of other foreign powers, and finally the series of
political events that drew a line in the sand for the colonies.

\subsubsection{Previous Wars and the Influence of Foreign Powers}

The American Revolution, from the international standpoint, was one in a series
of conflicts between Europe's major powers that also took part in America.
Britain and France had already been at war off an on for 74 years before
the Revolution, during the Intercolonial Wars. Table \ref{intercolonial_wars})
describes these conflicts.

%%
% Table \ref{intercolonial_wars} describes these conflicts. Each of the first
% three intercolonial wars left the question of borders between French and
% British colonies open, motivating further conflict. The first three wars
% started in Europe, but the French and Indian War began in America when the
% French claimed areas settled by the English colonists.
%%

\input{22-intercolonial_wars}
Each of the first three Intercolonial wars left the question of borders between
French and British open, motivating further conflict. The first three wars
started in Europe, however the French and Indian War began in the colonies in
1754 when the French claimed areas settled by the English colonists.

In 1756, Frederick II, the Great, of Prussia conducted a pre-emptive invasion of
Saxony. Austria, France, Russia, Sweden and Saxony allied against Prussia and
Britain. With this attack, Frederick ``succeeded only in solidifying the
coalition against him,''\cite{dupuy_1977} and drew Britain as an ally into a
general conflict between European powers. 

Ultimately the Prussians were saved by the favor of Peter III when he succeeded
Empress Elizabeth to the throne of Russia in 1762, rather than by the British,
despite their support. As Dupuy describes, the European land campaigns of
the Seven Years' War, fought over dynastic aims, had little lasting impact.
However, the naval and foreign campaigns fought between Britain and France over
empire and mercantilism, reshaped the sea-lanes in Britain's favor.

The Seven Years' War allowed Britain to route the French from
%nearly all of North America except New Orleans and the North Atlantic fisheries
Canada, all lands east of the Mississippi River, India, several Caribbean
islands,\footnote{``In the Caribbean, British forces captured two French islands
valuable for their sugar, Guadeloupe and Martinique, territories considered by
many more crucial to the French Empire than all of vast, vacant Canada. They
also took Grenada, St. Lucia, St. Vincent, and Dominica.'' \cite[8]{dupuy_1977}}
Gor\'ee, France's chief slave-trading center in Western Africa, as well as
Gibraltar and Minorca.\cite[7-9]{dupuy_1977} From Spain, Britain seized Havana,
Manila and the Philippines.

George Grenville, Pitt's political successor after the assession of King George III in
Oct. 1760, saw the British debit at \pounds140 millon with \pounds5 million
annual interest, and was determined to reduce it. To accomplish this, he severely
weakened the Navy, tempting both France and Spain, and began to levy `internal'
taxes on the colonies, which was a break with all previous policy, and a severe
grevance when all appeals from colonial legislatures were
ignored.\cite[51-59]{cook_long_1995}

After a whole generation of war over territories and trade, France and Spain
would not easily forget Britain's gains during the previous war, and were ready
to exploit any weakness. As William Pitt, Secretary of State in Britain
during the latter portion of the war was well aware. He and others were
concerned that any perceived weakness within Britain would give France and
Spain another opening to attack, especially in it's weakened economic
and military strength after the war. William Pitt was relieved of power by the
king, succeeded by a procession four other prime ministers with weak political power or
flawed policies in only five years. Finally, after the failure of Grenville's
policies, Pitt was finally re-instated by the king as a background leader behind
a young duke of Grafton and Charles Townshend as chancellor fo the Exchequer.
Although the colonies were thrilled with Pitt's return to power, even erecting
statues of him, Pitt was too ill to lead, and Townshend took the political lead.

King George III ascended to the throne in 1760, during the Seven Years War, and
was in open conflict with both William Pitt, despite his strategic leadership
which yielded victories for the British. Rather than . He succeeded two German speaking
kings, and was the first native-born sovereign over England in two generations.
Along with these other differences, he was at odds with the Whig party which has
previously driven politics and policy in England for during the reigns of George
I and II. His transition was three years prior to the end of the Seven Years'
War, although crowned following a series of victories, George III's first
proclamation as ruler reflected his desire to end the war, rebuild a
war-burdened economy and country that had already reached it's highest level of
debit so far.

\begin{quote}
``George III was twenty-two years old when he began a history-laden reign that
would last for sixty years. He was headstrong and obstinate, full of youthful
prejudices, lacking in humility, wisdom, or vision, with little interest in the
opinions of others. But he was also hardworking, prudish, religious, and moral.
He came to the throne determined not merely to reign but to rule. To do so, he
would have to bring his own Loyalists to power and open a new political era
after a half century of dominance by the Whigs. He judged that England was
war-weary and overburdened with debt. He wanted peace, and he took no great
pride in Pitt's faraway victories in the American colonies or other parts of the
globe. American trade was booming, there were no signs of trouble, and all
seemed well.'' \cite[p. 5]{cook_long_1995}
\end{quote}

%\cite[10]{dupuy_1977}
%
%\subsubsection{Mercantilism}
%
%%\subsubsection{Foreign Powers}
%
%\subsubsection{The Political Path to War}

%\begin{quote}
%``A great Empire and little minds go ill together,'' Edmund Burke in the House
%of Commons. March 1775
%\end{quote}


``Historians generally contend that once the British Army had cleared the French
out of North America and the colonies were no longer threatened by this outside
power, American independence became inevitable.\ldots'' although Cook explains
that this was not inevitable when George III first began his reign, before the
end of `Salutory Neglect' and the increased direct rule.\cite[p. 5]{cook_long_1995} 

%Expansion of Empire during/after the Seven Years' War
%
%\begin{quote}\red{
%``At the end of the Seven Year's War, her merchant fleet totaled some 500,000
%tons of sailing ships.''\cite[p. 15]{cook_long_1995}
%}\end{quote}

The political and economic origins of the American revolution are tied to King
George III's resentment of the Whig parliament that had ruled England during the
reign of his father and grandfather. George III came to the throne as a native
Brit, rather than a German import, and rather than allowing parliament and the
Prime Minister to govern, he was determined to replace the Whigs with Loyalists
and restore power to the throne. The political maneuvering of his followers,
the leadership of William Pitt, the popular Whig secretary of state and
victorious strategist of the Seven Years War, and the `salutary
neglect'  policies of his predecessor Sir Robert Walpole, Prime Minister under
King George the I.\cite[Ch 1-4]{cook_long_1995}

Walpole ``took the enlightened view that if the colonies were left to run their
own local affairs with minimal interference from London, they would produce more
wealth and commerce, prosper, and give less trouble.''\cite[p. 3]{cook_long_1995}

This policy was not allowed to continue during George III's reign however, as a
procession of prime ministers and other leadership starting and \red{Pitt} and,
ironically, effectively ending with \red{Pitt's} illness and the rise of Charles
Townshend who committed the final offenses towards any positive policies towards
the colonies, eroding and finally destroying `salutary neglect' and
any chance of recovering Colonial support for British taxation and military
presence in the Colonies.

%George Grenville
%
%\ldots
%
%Charles Townshend
%\footnote{
%  February 6, 1765:
%
%  ``They, planted by your care? No! Your oppression planted 'Em in America,
%  They fled your tyranny to an uncultivated an unhospitable country where they
%  exposed themselves to almost all the hardships to which human nature is
%  liable, and among others to the cruelty of a Savage foe. And yet, actuated by
%  the Principles of English Liberty, they met all these hardships with pleasure,
%  compared with those they suffered in their own Country, from the hands of
%  those who should have been their friends.''
%
%  ``They, nourished by your Indulgence? They grew up by your neglect of 'Em; as
%  soon as you began to care about 'Em---sent to spy out their Liberty, to
%  misrepresent their Actions, to prey on 'Em.''
%
%  ``They, protected by your Arms? They have nobly taken up Arms in your defense,
%  have exerted a valour against their consent and Laborious industry for the
%  defense of a Country whose frontiers, while drenched in blood, its interior
%  Parts have yielded all its little Savings to your Emolument. And believe me,
%  and remember I this day told you so, that some spirit of Freedom which
%  actuated the at people at first will accompany them still\ldots
%  
%  ''I claim to know more of America that most of you, having seen and been
%  conversant in that Country. The People I believe are as truly Loyal as any
%  Subjects the King has, but a people Jealous of their Liberties who will
%  vindicate them, if ever they should be violated\ldots''
%  
%  Colonel Isaac Barr\'e during debates in the House of Commons on the Stamp act
%  on Feb. 6, 1765, in reply to Charles Townshend's rhetoric describing the
%  Americans as ``children planted by our care, nourished by our indulgence\ldots
%  protected by our arms'' who were begrudging England their rightful returns for
%  protection and defense of the colonies.\cite[p. 67]{cook_long_1995}
%} 

%\ldots
%
%\ldots

\input{12-timeline}


%%% --begin-notes
%\subsection{--notes--}
%
%\footcite[2]{moncure_cowpens_1996}
%
%Enlightenment political theory:
%
%Village green in Lexington, MA 19 APR 1775
%
%trade unrestricted prior to 1760s
%
%King sends:\\
%Major General Sir William Howe\\
%Major General Sir Henry Clinton\\
%Major General Sir John Burgoyne
%
%\paragraph{--notes--}\footcite[]{}
%``Shy. John, \emph{A People Numerous and Armed.} pp. 198-199 and 230-232. The
%British contemplated cutting loose New England and retaining the richer lands of
%the South in 1778. but never pursued a negot~ated settlement with the
%Continental Congress.''

%\input{13-brit_table.tex}
