% vim: set nowrap: set tw=999
\begin{singlespacing}
{\small
\ctable[
	cap={The Intercolonial Wars},
	caption={The Intercolonial Wars (1689-1763)\tmark[a]},
	pos=h,
	label={intercolonial_wars}]
	{lllp{1.4in}p{1.8in}p{1.6in}}{
	\tnote[a]{\fullcite[98-111]{berard_1878} and table adapted from 
	\fullcite{wiki:fr_ind_2011}}
	}{\FL
War & Start & End  & As known\ \newline in the Colonies:   & \ \newline in Europe: 									& Settled by:				\ML
1st & 1689  & 1697 & King William's War		           & War of the Grand Alliance			&	Treaty of Ryswick		\NN
    & \multicolumn{5}{p{5.8in}}{\footnotesize
	King William III succeeds James II. Louis XIV of France declares war on
	England in support of James II. French also attack New York and New
	Hampshire. Colonies receive no support from England and organize mutual
	defense. Treaty of Ryswick does not set clear boundaries between French
	and English colonies. \cite[98-99]{berard_1878}
}\NN
2nd & 1702  & 1713 & Queen Anne's War		           & War of the Spanish Succession		&	Treaty of Utrecht		\NN
    & \multicolumn{5}{p{5.8in}}{\footnotesize
    	Originated in Europe. Spain and France vs. England. South Carolina sent
	and expedition into Florida, strengthening England's claim on Georgia.
	French encouraged Indians to attack New England colonies, prompting more
	inter-colonial coordination and a bounty on indian scalps. Annapolis
	captured and England sent a failed invasion into Canada. The treaty
	required France to yield some areas in Nova
	Scotia.\cite[100-101]{berard_1878}
}\NN
3rd & 1744  & 1748 & King George's War		           & War of Jenkins' Ear/ Austrian Succession	&	Treaty of Aix-la-Chapelle	\NN
    & \multicolumn{5}{p{5.8in}}{\footnotesize
    	The French begin colonial hostilities by recapturing areas in Nova
	Scotia. Multiple colonies coordinate to capture Louisburg from the
	French after a fifty day siege. Louisburg was restored to the French in
	the treaty of Aix-la-Chapelle. Colonial boundaries are still
	undefined.\cite[101-103]{berard_1878}
}\NN
4th & 1754  & 1763 & French and Indian War	           & Seven Years' War &	Treaty of Paris			\NN
    & \multicolumn{5}{p{5.8in}}{\footnotesize
    	Only six years after King George's War, the French claim the Mississippi
	Valley and all tributary valleys, prompting a conflict over the Ohio
	River Valley, which the British determine to colonize. France was
	largely successful until 1758, when the new prime minister \red{Pitt}
	changed strategy and policies. France was routed from all of North
	America, minus the fisheries, ceding it's possessions to England and
	Spain.\cite[103-111]{berard_1878}
}\LL
}}
\end{singlespacing}
%\footnote{\url{http://en.wikipedia.org/wiki/French_and_Indian_Wars}}
%
%  The War of Grand Alliance and King Williams War 
%  
%  In the War of Austrian Succession (1740-1748) Britain and Austria allied against
%  France, Spain and others, while in the Americas, Britan's colonies fought the
%  French and Indians in King George's War (1744-1748) \footnote{Rickard, J. (22
%  October 2000), King George's War (1744-18 October 1748),
%  \url{http://www.historyofwar.org/articles/wars_kinggeorge.html}}
